Robust-Adaptive Control of Linear Systems:\\ beyond Quadratic Costs

### 2020-10-12 main.tex
Add paragraph and acks, final option

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2020
\PassOptionsToPackage{numbers, compress}{natbib}
% ready for submission
 \usepackage{neurips_2020}

@@

% to compile a camera-ready version, add the [final] option, e.g.:
%     \usepackage[final]{neurips_2020}

@@

\author{%
	David S.~Hippocampus\thanks{Use footnote for providing further information
		about author (webpage, alternative address)---\emph{not} for acknowledging
		funding agencies.} \\
	Department of Computer Science\\
	Cranberry-Lemon University\\
	Pittsburgh, PA 15213 \\
	\texttt{hippo@cs.cranberry-lemon.edu} \\
	% examples of more authors
	% \And
	% Coauthor \\
	% Affiliation \\
	% Address \\
	% \texttt{email} \\
	% \AND
	% Coauthor \\
	% Affiliation \\
	% Address \\
	% \texttt{email} \\
	% \And
	% Coauthor \\
	% Affiliation \\
	% Address \\
	% \texttt{email} \\
	% \And
	% Coauthor \\
	% Affiliation \\
	% Address \\
	% \texttt{email} \\
}

@@

The control of uncertain systems is a long-standing problem, to which a vast body of literature is dedicated. Existing work is mostly concerned with the problem of \emph{stabilisation} around a fixed reference state or trajectory, including approaches such as $\cH_\infty$ control \citep[][]{Basar1996}, sliding-mode control \citep{Lu1997} or system-level synthesis \citep{Dean2017,Dean2018}. This paper fits in the popular MPC framework, for which adaptive data-driven schemes have been developed to deal with model uncertainty \citep{Sastry1990,Tanaskovic2014,Amos2018}, but lack guarantees. The family of tube-MPC algorithms seeks to derive theoretical guarantees of \emph{robust constraint satisfaction}: the state $x$ is constrained in a safe region $\mathbb{X}$ around the origin, often chosen convex \citep{Fukushima2007,Adetola2009,Aswani2013,Turchetta2016,Lorenzen2017,Kohler2019,Lu2019}. %, their adaptive versions take also into account the possible parametric model uncertainty
Yet, many tasks cannot be framed as stabilisation problems (e.g. obstacle avoidance) and are better addressed with the minimax control objective, which can allow more flexible goal formulations. Minimax control has mostly been studied in two particular instances.

@@

This bound contains an input-dependent \hlb{estimation error} term, which can be further bounded under an additional assumption that the features are sufficiently excited.

@@

The procedure we developed in sections \ref{sec:estimation}, \ref{sec:prediction} and \ref{sec:control} relies on strong modelling assumptions, such as the linear dynamics \eqref{eq:dynamics} and \Cref{assumpt:structure}. But what if they are wrong?

\paragraph{Model adequacy}
One of the major benefits of using the family of linear models, compared to richer model classes, is that they provide strict conditions allowing to quantify the adequacy of the modelling assumptions to the observations.
Given $N-1$ observations, \Cref{sec:estimation} provides a polytopic confidence region \eqref{eq:polytope} that contains $A(\theta)$ with probability at least $1-\delta$. Since the dynamics are linear, we can propagate this confidence region to the next observation: $y_{N}$ must belong to the Minkowski sum of a polytope representing model uncertainty $\cP(A_{0} x_N + Bu_N, \Delta A_{1}x_N,\dots, \Delta A_{2^d}x_N)$ and a polytope $\cP(0_p, \underline{\eta}, \overline{\eta})$ bounding the disturbance and measurement noises. \citet{delos2015} provide a way to test this membership in polynomial time using linear programming. Whenever it is not verified, we can confidently reject the $(A,\phi)$-modelling \Cref{assumpt:structure}. This enables us to consider a rich set of potential features $\left((A^1, \phi^1), \dots, (A^M, \phi^M)\right)$ rather than relying on a single assumption, and only retain those that are consistent with the observations so far. Then, every remaining hypothesis must be considered during planning.

@@

\begin{assumption}[Multi-model ambiguity]
\label{assumpt:multi-model-ambiguity}
The true dynamics $f$ lies within a finite set of candidate models $f^1, \dots, f^M$:
$$
\exists m\in[M]: \dot{x}(t) = f^m(x(t), u(t)), \forall t\geq 0.
$$
\end{assumption}
We adapt our planning algorithm to balance these concurrent hypotheses in a robust fashion, i.e. maximise a robust objective with discrete ambiguity:
\begin{equation}
\label{eq:robust-objective-discrete}
V^r = \sup_{a\in\cA^\Natural}\min_{m\in[M]} \sum_{n=N+1}^\infty \gamma^n R_n^m, 
%\refstepcounter{equation}\hfill(\theequation)\label{eq:robust-objective-discrete}
\end{equation}
where $R_n^m$ is the reward obtained by following the action sequence $a$ up to step $n$ under the dynamics $f^m$.
This objective could be optimised in the same way as in \Cref{sec:control}, but this would result in a coarse and lossy approximation. Instead, we exploit the finite uncertainty structure of \Cref{assumpt:multi-model-ambiguity} to asymptotically recover the true $V^r$ by modifying the \texttt{OPD} algorithm in the following way:

\begin{definition}[Robust sequence upper bounds] We replace the upper-bound \eqref{eq:opd} on sequence values in \texttt{OPD} by:
$$
%\label{eq:robust-b-values}
B_a^r(k)  \eqdef \min_{m\in[M]} \sum_{n=0}^{h-1} \gamma^n R_n^m  + \frac{\gamma^h}{1-\gamma}. \refstepcounter{equation}\hfill(\theequation)\label{eq:robust-b-values}
$$
\end{definition}
Note that it is not equivalent to solving each control problem independently and following the action with highest worst-case value, as we show in the Supplementary Material. We analyse the sample complexity of the corresponding robust planning algorithm.

@@

\paragraph{Motion planning for an autonomous vehicle}
We consider the \href{https://github.com/eleurent/highway-env}{highway-env} environment \citep{highway-env} for simulated driving decision problems. An autonomous vehicle with state $\chi_0\in\Real^4$ is approaching an intersection among $V$ other vehicles with states $\chi_i\in\Real^4$, resulting in a joint traffic state $x = [\chi_0, \dots,\chi_V]^\top\in\Real^{4V+4}$. These vehicles follow parametrized behaviours $\dot{\chi}_i=f_i(x,\theta_i)$ with unknown parameters $\theta_i\in\Real^5$. We appreciate a first advantage of the structure imposed in \Cref{assumpt:structure}: the uncertainty space of $\theta$ is $\Real^{5V}$. In comparison, the traditional LQ setting where the whole state matrix $A$ is estimated would have resulted in a much larger parameter space $\theta\in\Real^{16V^2}$.
%This allows to scale to larger systems: in our experiments, we used a state space of dimension $44$ ($V=10$) where e.g. \citep{Dean2018,abeille18a} reported numerical experiments with states of dimensions 3 and 4, respectively.
The system dynamics $f$, which describes the interactions between vehicles, can only be expressed in the form of \Cref{assumpt:structure} given the knowledge of the desired route for each vehicle, with features $\phi$ expressing deviations to the centerline of the followed lane. Since these intentions are unknown to the agent, we adopt the multi-model perspective of \Cref{sec:multi-model} and consider one model per possible route for every observed vehicle before an intersection. In \Cref{tab:driving}, we compare \Cref{alg:full} to a nominal agent planning with the estimated parameter $\theta_{N,\lambda}$, with two different modelling assumptions: we assume that Nominal 1 has access to the true followed route for each vehicle, while Nominal 2 does not and picks the model with minimal prediction error. As before, the robust approach is less efficient on average than planning with a nominal model, but it does better in terms of worst-case performance and manages to avoid collisions, contrary to the nominal baselines that suffer from model bias. Again, we also compare to the final policy of a DQN baseline trained over 3000 episodes, causing $1058\pm113$ collisions while training ($35\pm4\%$). We provide more details about the system and videos of the experiments in the Supplementary Material.

@@

>>>>

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2020
\PassOptionsToPackage{numbers, compress}{natbib}
% ready for submission
% \usepackage{neurips_2020}

@@

% to compile a camera-ready version, add the [final] option, e.g.:
     \usepackage[final]{neurips_2020}

@@

\author{%
	Edouard~Leurent \\
	Inria Lille Nord-Europe \& Renault\\
	Paris, France\\
	\texttt{edouard.leurent@inria.fr} \\
	% examples of more authors
	 \And
	 Denis~Efimov \\
	 Inria Lille Nord-Europe, Valse team\\
	 Lille, France\\
	 \texttt{denis.efimov@inria.fr} \\
	 \And
	 Odalric-Ambrym~Maillard \\
	 Inria Lille Nord-Europe, SequeL team\\
	 Lille, France\\
	 \texttt{odalric.maillard@inria.fr} \\
}

@@

The control of uncertain systems is a long-standing problem, to which a vast body of literature is dedicated. Existing work is mostly concerned with the problem of \emph{stabilisation} around a fixed reference state or trajectory, including approaches such as $\cH_\infty$ control \citep[][]{Basar1996}, sliding-mode control \citep{Lu1997} or system-level synthesis \citep{Dean2017,Dean2018}. This paper fits in the popular MPC framework, for which adaptive data-driven schemes have been developed to deal with model uncertainty \citep{Sastry1990,Tanaskovic2014,Amos2018}, but lack guarantees. The family of tube-MPC algorithms seeks to derive theoretical guarantees of \emph{robust constraint satisfaction}: the state $x$ is constrained in a safe region $\mathbb{X}$ around the origin, often chosen convex \citep{Fukushima2007,Adetola2009,Aswani2013,Turchetta2016,Lorenzen2017,Kohler2019,Lu2019,Leurent2020robust}.
Yet, many tasks cannot be framed as stabilisation problems (e.g. obstacle avoidance) and are better addressed with the minimax control objective, which can allow more flexible goal formulations. Minimax control has mostly been studied in two particular instances.

@@

It is difficult to check the validity of the stability condition 2. since it applies to matrices $A_N$ produced by the algorithm rather than to the system parameters. A stronger but easier to check condition is that the polytope \eqref{eq:polytope} at some point becomes included in a set where this property is uniformly satisfied. For instance, if the features are sufficiently excited, the estimation converges to a neighbourhood of the true dynamics $A(\theta)$. This also allows to further bound the input-dependent \hlb{estimation error} term.

@@

The procedure we developed in \Cref{sec:estimation,sec:prediction,sec:control} relies on strong modelling assumptions, such as the linear dynamics \eqref{eq:dynamics} and \Cref{assumpt:structure}. But what if they are wrong?

\paragraph{Model adequacy}
One of the major benefits of using the family of linear models, compared to richer model classes, is that they provide strict conditions allowing to quantify the adequacy of the modelling assumptions to the observations.
Given $N-1$ observations, \Cref{sec:estimation} provides a polytopic confidence region \eqref{eq:polytope} that contains $A(\theta)$ with probability at least $1-\delta$. Since the dynamics are linear, we can propagate this confidence region to the next observation: $y_{N}$ must belong to the Minkowski sum of a polytope representing model uncertainty $\cP(A_{0} x_N + Bu_N, \Delta A_{1}x_N,\dots, \Delta A_{2^d}x_N)$ and a polytope $\cP(0_p, \underline{\eta}, \overline{\eta})$ bounding the disturbance and measurement noises. \citet{delos2015} provide a way to test this membership in polynomial time using linear programming. Whenever it is not verified, we can confidently reject the $(A,\phi)$-modelling \cref{assumpt:structure}. This enables us to consider a rich set of potential features $\left((A^1, \phi^1), \dots, (A^M, \phi^M)\right)$ rather than relying on a single assumption, and only retain those that are consistent with the observations so far. Then, every remaining hypothesis must be considered during planning.

@@

\begin{assumption}[Multi-model ambiguity]
\label{assumpt:multi-model-ambiguity}
The dynamics $f$ lie in a finite set of candidates $(f^m)_{m\in[M]}$.
%$
%\exists m\in[M]: \dot{x}(t) = f^m(x(t), u(t)), \forall t\geq 0.
%$
\end{assumption}
We adapt our planning algorithm to balance these concurrent hypotheses in a robust fashion, i.e. maximise a robust objective with discrete ambiguity:
\begin{equation}
\label{eq:robust-objective-discrete}
V^r = \sup_{a\in\cA^\Natural}\min_{m\in[M]} \sum_{n=N+1}^\infty \gamma^n R_n^m, 
%\refstepcounter{equation}\hfill(\theequation)\label{eq:robust-objective-discrete}
\end{equation}
where $R_n^m$ is the reward obtained by following the action sequence $a$ up to step $n$ under the dynamics $f^m$.
This objective could be optimised in the same way as in \Cref{sec:control}, but this would result in a coarse and lossy approximation. Instead, we exploit the finite uncertainty structure of \Cref{assumpt:multi-model-ambiguity} to asymptotically recover the true $V^r$ by modifying the \texttt{OPD} algorithm in the following way:

\begin{definition}[Robust UCB] We replace the upper-bound \eqref{eq:opd} on sequence values in \texttt{OPD} by:
\begin{equation}
\label{eq:robust-b-values}
B_a^r(k)  \eqdef \min_{m\in[M]} \sum_{n=0}^{h-1} \gamma^n R_n^m  + \frac{\gamma^h}{1-\gamma}. 
\end{equation}
\end{definition}
Note that it is not equivalent to solving each control problem independently and following the action with highest worst-case value, as we show in the Supplementary Material. We analyse the sample complexity of the corresponding robust planning algorithm.

@@

\textbf{Motion planning for an autonomous vehicle}~~
We consider the \href{https://github.com/eleurent/highway-env}{highway-env} environment \citep{highway-env} for simulated driving decision problems. An autonomous vehicle with state $\chi_0\in\Real^4$ is approaching an intersection among $V$ other vehicles with states $\chi_i\in\Real^4$, resulting in a joint traffic state $x = [\chi_0, \dots,\chi_V]^\top\in\Real^{4V+4}$. These vehicles follow parametrized behaviours $\dot{\chi}_i=f_i(x,\theta_i)$ with unknown parameters $\theta_i\in\Real^5$. We appreciate a first advantage of the structure imposed in \Cref{assumpt:structure}: the uncertainty space of $\theta$ is $\Real^{5V}$. In comparison, the traditional LQ setting where the whole state matrix $A$ is estimated would have resulted in a much larger parameter space $\theta\in\Real^{16V^2}$.
%This allows to scale to larger systems: in our experiments, we used a state space of dimension $44$ ($V=10$) where e.g. \citep{Dean2018,abeille18a} reported numerical experiments with states of dimensions 3 and 4, respectively.
The system dynamics $f$, which describes the interactions between vehicles, can only be expressed in the form of \Cref{assumpt:structure} given the knowledge of the desired route for each vehicle, with features $\phi$ expressing deviations to the centerline of the followed lane. Since these intentions are unknown to the agent, we adopt the multi-model perspective of \Cref{sec:multi-model} and consider one model per possible route for every observed vehicle before an intersection. In \Cref{tab:driving}, we compare \Cref{alg:full} to a nominal agent planning with two different modelling assumptions: Nominal 1 has access to the true followed route for each vehicle, while Nominal 2 does not and picks the model with minimal prediction error. Again we also compare to a DQN baseline trained over 3000 episodes, causing $1058\pm113$ collisions while training ($35\pm4\%$). We observe similar results as before: the robust agents has a higher worst-case performance and avoids collisions at all time, at the price of a decreased average performance. We provide more details about the system and videos of the experiments in the Supplementary Material.

@@

\begin{ack}
	This work was supported by the French Ministry of Higher Education and Research, and CPER Nord-Pas de Calais/FEDER DATA Advanced data science and technologies 2015-2020.
\end{ack}
